Harnessing AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology

### 2024-04-18 _manuscript/manuscript.qmd
add manuscript files

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)
library(HPOExplorer)
```

@@

```{r occurence_plot}
#| echo: false
#| warning: false

@@

# Stacked bar plot of clinical characteristic occurrence
ggplot(count_df_long, aes(x = category, y = count, fill = condition)) +
  geom_bar(stat = "identity") +
  labs(title = "Clinical characteristic occurrence",
       x = "", y = "HPO phenotypes (n)", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "right", plot.title = element_text(hjust=0.5), 
        text = element_text(size=16),
        axis.text.x = element_text(angle=45, size=16, hjust=1), 
        title = element_text(size=15)) +
  scale_fill_brewer(palette = "GnBu", direction = -1)

@@

```{r annot_checks}
# Annotation checks
checks <- HPOExplorer::gpt_annot_check()

@@

```{r figure_2}
#| label: fig-checks
#| fig-cap: GPT-4 annotations are consistent and accurate across annotations. a Barplot showing the annotation consistency for 793 phenotypes that were annotated twice. On the left hand side, annotations were defined as consistent if the given phenotype was annotated to the same outcome occurrence both times, or if it was annotated to ‘always’ and ‘often’, or, ‘never’ and ‘rarely. On the right hand side, annotations were only defined as consistent if the given phenotype was annotated to the same outcome occurrence both times. b Bar plot of the true positive rate for each annotation. The true positives were determined based on the assumption that all descendants of some HPO phenotypes should be related to those outcomes. For instance, all descendents of the HPO term for “Neoplasm” were considered to be true positives for cancer (an assumption which is not always the case, as GPT-4 detects, as some neoplasms are benign). The labels above each bar indicate the number of phenotypes tested. 
#| fig-height: 4
#| fig-width: 6
consistency_plot <- ggplot(check_df[metric_type %in% metric_types & metric_category %in% metric_categories, ],
                           aes(x = annotation, y = value,
                               fill = metric_category,
                               label = round(value, 2))) +
  geom_bar(stat = "identity", show.legend = TRUE) +
  facet_grid(facets = ~metric_category, scales = "free") +
  theme_bw() +
  labs(x = "", y = "Value") +
  ggtitle("Annotation consistency") +
  scale_fill_manual(values = c("#C3DAEAFF", "#ECE28BFF")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
        legend.position = "none",
        text = element_text(size = 16),
        title = element_text(size = 15),
        plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(x = 7.5, y = 1.05, label = "N phenotypes = 793"), size = 4)

@@

```{r severity_scoring}
# get all descendants of the term phenotypic abnormality 
keep_descendants="Phenotypic abnormality"
keep_ont_levels=seq(3,17)

@@

```{r figure_3}
#| label: fig-severity
#| fig-cap: Quantifying the severity of HPO phenotype annotations highlights the most impactful conditions. Heatmap of the top (n = 15), mid-range (n = 15), and bottom (n = 15) HPO phenotypes based on severity score. Severity scores were calculated by multiplying the numeric values assigned to each clinical characteristic (death = 5, intellectual disability = 5, impaired mobility = 4, blindness = 4, congenital onset = 4, physical malformations = 3, sensory impairments = 3, immunodeficiency = 3,  cancer = 3, reduced fertility = 1) by their corresponding response frequencies (Always = 4, Often = 3, Rarely = 2, Never = 0). The average normalised score, representing overall phenotype severity on a 0-100 scale, was calculated by aggregating the multiplied values and normalising by the theoretical maximum severity score. The x-axis is scaled by the severity of the clinical characterisation.
#| fig-height: 4
#| fig-width: 6

@@

| Clinical characteristic     | Always (4) | Often (3) | Rarely (2) | Never (0) |
| Intellectual_disability (5) |     20     |     15    |    10      |     0     | 
| Impaired_mobility (5)       |     20     |     15    |    10      |     0     | 
| Blindness (4)               |     16     |     12    |    8       |     0     | 
| Congenital onset (4)        |     16     |     12    |    8       |     0     | 
| Physical_malformations (3)  |     12     |     9     |    6       |     0     | 
| Sensory_impairments (3)     |     12     |     9     |    6       |     0     | 
| Immunodeficiency (3)        |     12     |     9     |    6       |     0     | 
| Cancer (3)                  |     12     |     9     |    6       |     0     | 
| Reduced_fertility (1)       |     4      |     3     |    2       |     0     |

@@

```{r supp_figure_1}
dat2 <- HPOExplorer:::gpt_annot_melt(res_coded = res_coded)
dat2 <- add_ancestor(dat2,
                     keep_descendants = keep_descendants)
dat2[,mean_severity_score_gpt:=mean(severity_score_gpt, na.rm=TRUE),
     by="ancestor_name"] |>
  data.table::setorderv("mean_severity_score_gpt", -1, na.last = TRUE)
dat2[,ancestor_name:=factor(ancestor_name,
                            levels = unique(dat2$ancestor_name),
                            ordered = TRUE)]

>>>>

```{r setup}
#| echo: false
#| warning: false

library(dplyr)
library(ggplot2)
library(patchwork)
library(HPOExplorer)
```

@@

```{r occurence_plot}
#| echo: false
#| warning: false
#| message: false

@@

# Stacked bar plot of clinical characteristic occurrence
figure1 <- ggplot(count_df_long, aes(x = category, y = count, fill = condition)) +
  geom_bar(stat = "identity") +
  labs(title = "Clinical characteristic occurrence",
       x = "", y = "HPO phenotypes (n)", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "right", plot.title = element_text(hjust=0.5), 
        text = element_text(size=16),
        axis.text.x = element_text(angle=45, size=16, hjust=1), 
        title = element_text(size=15)) +
  scale_fill_brewer(palette = "GnBu", direction = -1)

@@

```{r annot_checks}
#| echo: false
#| message: false

# Annotation checks
checks <- HPOExplorer::gpt_annot_check()

@@

```{r figure_2}
#| echo: false
#| message: false
#| label: fig-checks
#| fig-cap: GPT-4 annotations are consistent and accurate across annotations. a Barplot showing the annotation consistency for 793 phenotypes that were annotated twice. On the left hand side, annotations were defined as consistent if the given phenotype was annotated to the same outcome occurrence both times, or if it was annotated to ‘always’ and ‘often’, or, ‘never’ and ‘rarely. On the right hand side, annotations were only defined as consistent if the given phenotype was annotated to the same outcome occurrence both times. b Bar plot of the true positive rate for each annotation. The true positives were determined based on the assumption that all descendants of some HPO phenotypes should be related to those outcomes. For instance, all descendents of the HPO term for “Neoplasm” were considered to be true positives for cancer (an assumption which is not always the case, as GPT-4 detects, as some neoplasms are benign). The labels above each bar indicate the number of phenotypes tested. 
#| fig-height: 7
#| fig-width: 10
consistency_plot <- ggplot(check_df[metric_type %in% metric_types & metric_category %in% metric_categories, ],
                           aes(x = annotation, y = value,
                               fill = metric_category,
                               label = round(value, 2))) +
  geom_bar(stat = "identity", show.legend = TRUE) +
  facet_grid(facets = ~metric_category, scales = "free") +
  theme_bw() +
  labs(x = "", y = "Value") +
  ggtitle("Annotation consistency") +
  scale_fill_manual(values = c("#C3DAEAFF", "#ECE28BFF")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
        legend.position = "none",
        text = element_text(size = 16),
        title = element_text(size = 15),
        plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(x = 7.5, y = 1.05, label = "N phenotypes = 793"), size = 4)

@@

```{r severity_scoring}
#| echo: false
#| message: false

# get all descendants of the term phenotypic abnormality 
keep_descendants="Phenotypic abnormality"
keep_ont_levels=seq(3,17)

@@

```{r figure_3}
#| echo: false
#| message: false
#| label: fig-severity
#| fig-cap: Quantifying the severity of HPO phenotype annotations highlights the most impactful conditions. Heatmap of the top (n = 15), mid-range (n = 15), and bottom (n = 15) HPO phenotypes based on severity score. Severity scores were calculated by multiplying the numeric values assigned to each clinical characteristic (death = 5, intellectual disability = 5, impaired mobility = 4, blindness = 4, congenital onset = 4, physical malformations = 3, sensory impairments = 3, immunodeficiency = 3,  cancer = 3, reduced fertility = 1) by their corresponding response frequencies (Always = 4, Often = 3, Rarely = 2, Never = 0). The average normalised score, representing overall phenotype severity on a 0-100 scale, was calculated by aggregating the multiplied values and normalising by the theoretical maximum severity score. The x-axis is scaled by the severity of the clinical characterisation.
#| fig-height: 11
#| fig-width: 10

@@

| Clinical characteristic     | Always (4) | Often (3) | Rarely (2) | Never (0) |
| --------------------------- | ---------- | --------- | ---------- | --------- |
| Intellectual_disability (5) |     20     |     15    |    10      |     0     | 
| Impaired_mobility (5)       |     20     |     15    |    10      |     0     | 
| Blindness (4)               |     16     |     12    |    8       |     0     | 
| Congenital onset (4)        |     16     |     12    |    8       |     0     | 
| Physical_malformations (3)  |     12     |     9     |    6       |     0     | 
| Sensory_impairments (3)     |     12     |     9     |    6       |     0     | 
| Immunodeficiency (3)        |     12     |     9     |    6       |     0     | 
| Cancer (3)                  |     12     |     9     |    6       |     0     | 
| Reduced_fertility (1)       |     4      |     3     |    2       |     0     |

@@

```{r supp_figure_1}
#| echo: false
#| message: false
#| fig-cap: The distribution of phenotypic severity scores across various branches of the HPO. The histogram plots show the frequency of severity scores for all phenotypes  descending from the specified HPO branches. The x-axis represents the severity scores, and the y-axis indicates the count of phenotypes corresponding to each score. The average severity score for each HPO branch is highlighted in red.
#| fig-height: 10
#| fig-width: 10
dat2 <- HPOExplorer:::gpt_annot_melt(res_coded = res_coded)
dat2 <- add_ancestor(dat2,
                     keep_descendants = keep_descendants)
dat2[,mean_severity_score_gpt:=mean(severity_score_gpt, na.rm=TRUE),
     by="ancestor_name"] |>
  data.table::setorderv("mean_severity_score_gpt", -1, na.last = TRUE)
dat2[,ancestor_name:=factor(ancestor_name,
                            levels = unique(dat2$ancestor_name),
                            ordered = TRUE)]
